\Title

### 2012-12-08 synmut.tex
some changes applied

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitch-hike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym}
\usepackage{color}
\usepackage{subfig}
\usepackage{graphicx}
%\usepackage[font=small, format=hang, labelfont={sf,bf}, figurename=Fig.]{caption}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\DeclareMathOperator\de{d\!}
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}

@@

\begin{abstract}
\noindent
Intrapatient HIV evolution is goverened by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach frequencies of order one.  We analyze time
series of viral sequences from the V1-C5 part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely fix in the
viral population. Simulations suggest that such synonymous mutations
have a (Malthisuan) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by linkage to neighbouring beneficial
nonsynonymous alleles (genetic draft). As far as the biological causes are
concerned, we detect a negative correlation between fixation of an allele and
its involvement in evolutionarily conserved RNA stem-loop structures.
This phenonenon is not observed in other parts of the HIV genome, in which
selective sweeps are less dense and the genetic architecture less constrained.
%In absence of antiretroviral treatment, HIV is very successful at producing
%mutants that are able to stay undetected by the host immune system for months,
%boosting the infection~\citep{richman_rapid_2003, bunnik_autologous_2008,
%moore_limited_2009}. The high mutation rate at the core of this process,
%however, also generates genetic noise in the background of beneficial alleles.
%Because of the limited ourcrossing of HIV, hitchhikers usually stay linked to
%the focal allele for a time comparable to its sweeping time, i.e. a few
%months~\citep{neher_recombination_2010, batorsky_estimate_2011}. The later fate
%of these accessory mutations depends more and more, as they decouple genetically
%from the escape allele via recombination, on their own fitness effects. In this
%study we show that, in a genetic region particularly dense of selective sweeps,
%synonymous hitchhikers tend to revert on a time scale of the order of 500 days,
%because of their deleterious fitness effects. In addition, we provide evidence
%that the biological origin for this deleterious effects resides, at least
%partially, in the disruption of macroevolutionarily conserved RNA secondary
%structures, termed ``insulating stems''.
\end{abstract}

@@

HIV evolves rapidly within a single host during the course of the infection. This evolution is driven by strong selection imposed by the host immune system via killer T cells (CTLs) and neutralizing antibodies (AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high mutation rate of HIV \citep{mansky}.
When the host develops a CTL or AB response against a particular viral epitope, mutations that reduce or prevent recognition of the epitope frequently emerge. Escape mutations in epitopes targeted by CTLs typically evolve early in infection and spread rapidly through the population \citep{McMichael}. Later in infection, the most rapidly evolving part of the HIV genome are the so called variable loops of the envelope proteins that need to avoid recognition by neutralizing anti bodies. Mutation in \env~ spread through the population within a few months (see \figurename~\ref{fig:aft}, solid lines). During chronic infection, the (Malthusian) effect size of this beneficial
mutations is of the order of $s_e \sim 0.01$~\citep{neher_recombination_2010}.

These escape mutations are selected for their effect on the amino acid sequence of the viral proteins. The viral genome, however, needs to meet additional constraints such as efficient processing and translation, nuclear export, and packaging into the viral capsid which operate on the RNA level. Purifying selection beyond the protein sequence is therefore expected, while it seems reasonable that the bulk of positive selection through the immune system should be restricted to amino acid sequences. A couple important RNA elements  are well characterized. For example, within \env{} a certain RNA sequence, called \rev{} response element (RRE), is used by HIV to enhance
nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
well studied case is the interaction between viral reverse transcriptase, viral
ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}. Recent studies
have shown that genetically engineered HIV strains with skewed codon usage bias
(CUB) patterns towards more or less abundant tRNAs replicate better or worse,
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.

@@

Despite evidence for functional importance of specific RNA sequences, synonymous mutations are commonly used as approximate neutral markers in studies of viral evolution. Neutral markers allow to make inference about the stochastic forces driving evolution \citep{smth}. Here, we characterize the dynamics of synonymous mutations in \env{} and show that a substantial fraction of these mutations are deleterious. The central quantity we investigate is the probability of fixation of a mutation, conditional on its population frequency. Even though the synonymous mutations are deleterious and cannot be used as neutral markers, we show the degree to which they hitch-hike with nearby non-synonymous mutations is very informative. Extending the analysis of fixation probabilities to the non-synonymous mutations, we show that time dependent selection or strong competition of escape mutations inside the same epitope are necessary to explain the observed patterns of fixation and loss.

One simple way to assess the neutrality of synonymous mutations is to look at
their level of conservation. Deleterious mutations at functional sites are
expected to be absent or rare across the viral population; vice versa, mutant
alleles that reach high frequencies are expected to be neutral. If genetic sites
are independent, the equilibrium frequency of a deleterious allele with fitness
$-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site per generation;
neutral alleles have no equilibrium frequency and can slowly fix via genetic
drift~\citep{ewens_mathematical_2004}.
t If the focal synonymous mutant is linked
to another nonneutral allele, however, its frequency is the result of the
combined fitness effects of both sites, and simple conservation-level analyses
fail. Since recombination in HIV is known to happen
rarely~\citep{neher_recombination_2010, batorsky_estimate_2011}, the genetic
context of the synonymous change at hand must be taken into account. Our
results underline the importance of the latter scenario for intrapatient HIV
evolution.

\section{Results}
A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to spread through the population and fix, while it is lost with probability $1-\nu$. This is a simple consequence of the fact that exactly one of present $N$ individuals will be the common ancestor of the entire population at a particular locus and this ancestor has a probability $\nu$ of carrying this mutations, see illustration in \FIG{illustration}. Deleterious or beneficial mutations, in contrast, should fix less or more often, respectively. Time series sequence data therefore suggest a simple way to investigate average properties of different classes of mutations.

\paragraph{Synonymous mutations in \env, C2-V5 are mostly deleterious}

\FIG{aft} shows time series data of the frequencies of all mutations observed \env, C2-V5, in patient ??\citep{shankarappa_consistent_1999,liu_selection_2006}. Despite many synonymous mutations reaching high frequency, very few fix. This observation in further quantified in panels ? and ?, that stratify the data of ?? patients (see methods) according to the frequency at which different mutations are observed. Considering all mutations in a frequency interval $\nu_0$ at some time $t_i$, we calculate the fraction that is found at frequency 1, frequency 0, or at intermediate frequency at later time points $t_f$. Plotting these fixed, lost, and polymorphic fraction against the time interval $t_f-t_i$, we see that most synonymous mutations segregate for roughly one year and are lost  much more frequently than expected. The ultimate probability of loss or fixation is shown as a function of the initial frequency $\nu_0$ in panel ??. In contrast to synonymous mutations, the non-synonymous seem to follow more a less the neutral expectation -- a point to which we will come back below.

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\includegraphics[width=\linewidth]{Shankarappa_fix_loss_dt_times}
\caption{Synonymous mutations rarely fix in \env, C2-V5: Panel A shows mutation 
frequency trajectories observed in patient ?? \cite{shankarappa_consistent_1999};  Nonsynonymous
and synonymous mutations are shown as solid and dashed lines, respectively. 
Colors indicate the position of the site along the C2-V5 region (red to blue) ADD COLORBAR. MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN SEPARATELY.
While non-synonymous mutations frequently fix, very few synonymous mutations do even though they are frequently observed at intermediate frequencies. Panel B quantifies time course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. The ultimate fraction of synonymous mutations that fix as a function of intermediate frequency $\nu_0$.
}
\label{fig:aft}
\end{center}
\end{figure}

\citet{bunnik_autologous_2008} present a longitudinal data sets on the entire \env~gene of ?? patients at ?? time points with ??-?? sequences each. Repeating the above analysis separately on the C2-V5 region studied above and the remainder of \env~ reveal strikingly different behavior inside and outside the hyper variable region. Within C2-V5, this data fully confirms the observations made in the data set by \citet{shankarappa_consistent_1999}. In the remainder of \env, however, observed synonymous mutations behave as if they were neutral; see \FIG{fixp}.

@@

These observations suggest that many of the synonymous mutations in the part of \env~that includes the hyper variable regions are deleterious, while outside this regions only roughly neutral mutations are polymorphic.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}}
\caption{Left panel: fixation probability of derived synonymous alleles is strongly
suppressed in C2-V5 versus other parts of the {\it env} gene, and of
nonsynonymous ones.
Right panel: especially hard is fixation of new alleles in conserved regions flanking the V
loops. The black dashed line is the prediction from neutral
theory, for comparison purposes. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

\paragraph{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
One possible {\it a priori} explanation for lack of fixation of synonymous mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA secondary structures are relevant for HIV replication,
mutations in nucleotides involved in those base pairs are expected  to be deleterious and to revert preferentially. 
Many functionally important secondary structure elements have been characterized, including  the  \rev{} response element (RRE) to enhance nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}. It has been suggested early on that parts of the viral genome that has the potential to form stems as better conserved that the remainder \citep{forsdyke_reciprocal_1995}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has been measured using the SHAPE assay (a biochemical reaction preferentially altering unpaired bases) \citep{watts_architecture_2009}. The SHAPE assay has shown that the variable regions V1 to V5 tend to be unpaired, while the conserved regions between those variable regions form stems. We partition all synonymous alleles observed
at intermediate frequencies above 10-15\% depending on their final destiny
(fixation or extinction). Subsequently, we align our sequences to the reference
NL4-3 strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
reactivities. As shown in \figurename~\ref{fig:SHAPE} (left panel) in a
cumulative histogram, the reactivity of fixed alleles are systematically larger
than of alleles that are doomed to extinction. In other words, alleles that are
likely to be breaking RNA helices are also more likely to revert and finally be
lost from the population. We then split the synonymous mutations in the C2-V5 region further into conserved and variable regions and found that the biggest depression in fixation probability is observed in the conserved stems, while the variable loops show little deviations from the neutral signature; see \FIG{SHAPE}B.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness effect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human housekeeping genes~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not found, however, evidence for any contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, although codon-optimized HIV seems to perform better
{\it in vitro}, the distance in CUB between HIV and human genes is not shrinking
at the macroevolutionary level. Second, within a single patient, we do not
observe any bias towards more human-like CUB in the synonymous mutations that
reach fixation rather than extinction. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
\subfloat{\includegraphics[width=0.49\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}}
\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
in vitro} chemical attack and shown that some nucleotides are more likely to
be involved in RNA secondary folds. C1-C5 regions, in particular, show
conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
among all derived alleles in those regions reaching frequencies of order one,
there is a negative correlation between fixation and involvement in a base
pairing in a RNA stem (left panel). The rest of the genome does not show any
correlation (right panel). There might be too few silent polymorphisms in the
first place, or the signal might be masked by non-functional RNA
structures. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

\paragraph{Deleterious mutations are brought to high frequency by hitch-hiking}
While the observation that some fraction of synonymous mutations is deleterious is not unexpected, it seems odd that we observe them at high population frequency -- at least in some regions of the genome. The region of \env~in which we observe deleterious mutations at high frequency is special in that it undergoes frequent adaptive changes to evade recognition by neutralizing antibodies \cite{Williamson}. Due to the limited amount of recombination in HIV \cite{neher_recombination_2010,batorsky_2011}, deleterious mutations that are linked to adaptive variants can reach high frequency \citep{maynard_smith}.

The potential for hitch-hiking is already apparent from the allele frequency trajectories in \FIG{aft}, where many mutations appear to change rapidly in frequency as a flock. Deleterious synonymous mutations can be amplified exponentially by selection on linked nonsynonymous sites, a process known as {\it genetic draft}~\citep{neher_genetic_2011}. In order to be advected to high frequency by a linked adaptive mutation, the deleterious effect of the mutation has to be substantially smaller than the adaptive effect. The latter was estimated to be on the order of $s_a = 0.01$ per day. The approximate magnitude of the deleterious effects can be estimated from \FIG{fixtimes}, that shows the distribution of times for synonymous alleles to reach the fix or get lost starting from intermediate frequencies. The typical time to loss is of the order of 500 days. If this loss is driven by the deleterious effect of the mutation, this corresponds to deleterious effects of roughly $0.002$ per day.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitch-hike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym}
\usepackage{color}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\DeclareMathOperator\de{d\!}
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}

@@

\begin{abstract}
\noindent
Intrapatient HIV evolution is goverened by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach frequencies of order one.  We analyze time
series of viral sequences from the V1-C5 part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely fix in the
viral population. Simulations suggest that such synonymous mutations
have a (Malthisuan) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by linkage to neighbouring beneficial
nonsynonymous alleles (genetic draft). As far as the biological causes are
concerned, we detect a negative correlation between fixation of an allele and
its involvement in evolutionarily conserved RNA stem-loop structures.
This phenonenon is not observed in other parts of the HIV genome, in which
selective sweeps are less dense and the genetic architecture less constrained.
\end{abstract}

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular viral epitope, mutations that reduce or prevent
recognition of the epitope frequently emerge. Escape mutations in epitopes
targeted by CTLs typically evolve early in infection and spread rapidly through
the population~\citep{mcmichael_immune_2009}. Later in infection, the most
rapidly evolving part of the HIV genome are the so called variable loops of the
envelope protein gp120, which need to avoid recognition by neutralizing ABs.
Mutations in \env~, the gene encoding for gp120, spread through the population
within a few months (see \figurename~\ref{fig:aft}, solid lines). During chronic
infection, the (Malthusian) effect size of this beneficial mutations is of the
order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.

These escape mutations are selected for their effect on the amino acid sequence
of the viral proteins. The viral genome, however, needs to meet additional
constraints such as efficient processing and translation, nuclear export, and
packaging into the viral capsid: all these processes operate at the RNA level. A
few important RNA elements are well characterized. For example, a certain RNA
sequence, called \rev{} response element (RRE), is used by HIV to enhance
nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
well studied case is the interaction between viral reverse transcriptase, viral
ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}. Furthermore, recent studies have shown that genetically
engineered HIV strains with skewed codon usage bias (CUB) patterns towards more
or less abundant tRNAs replicate better or worse,
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
Purifying selection beyond the protein sequence is therefore expected, while it
seems reasonable that the bulk of positive selection through the immune system
should be restricted to amino acid sequences.

@@

Despite evidence for functional importance of specific RNA sequences, synonymous
mutations are commonly used as approximate neutral markers in studies of viral
evolution. Neutral markers allow to make inference about the stochastic forces
driving evolution~\citep{smth}. Here, we characterize the dynamics of synonymous
mutations in \env{} and show that a substantial fraction of these mutations are
deleterious. The central quantity we investigate is the probability of fixation
of a mutation, conditional on its population frequency. Even though the
synonymous mutations are deleterious and cannot be used as neutral markers, we
show that the degree to which they hitchhike with nearby non-synonymous
mutations is very informative; their ability to hitchhike for extended times is
itself rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the non-synonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

%One simple way to assess the neutrality of synonymous mutations is to look at
%their level of conservation. Deleterious mutations at functional sites are
%expected to be absent or rare across the viral population; vice versa, mutant
%alleles that reach high frequencies are expected to be neutral. If genetic sites
%are independent, the equilibrium frequency of a deleterious allele with fitness
%$-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site per generation;
%neutral alleles have no equilibrium frequency and can slowly fix via genetic
%drift~\citep{ewens_mathematical_2004}. This intuitive picture does not hold in
%presence of genetic linkage and, in particular, for HIV evolution, because
%recombination is rare~\citep{neher_recombination_2010, batorsky_estimate_2011}.
%A more likely scenario, at least in escape mutation-rich regions, is the
%following: if the focal synonymous mutant is linked to a beneficial allele
%nearby, the latter essentially carries the synonymous allele toward high
%frequencies for a time on the order of the inverse of the recombination rate.
%The two slowly decouple afterwards, and the fitness effect of the synonymous
%allele starts to be visible on its own. On the one hand, genetic linkage and
%hitchhiking confound the interprentation of conservation levels; on the other,
%as we show below, purifying selection can be still observed with the help of
%longitudinal data.

\section{Results}

A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to
spread through the population and fix, while it is lost with probability
$1-\nu$. This is a simple consequence of the fact that exactly one of present
$N$ individuals will be the common ancestor of the entire population at a
particular locus and this ancestor has a probability $\nu$ of carrying this
mutations, see illustration in \FIG{fixp}. Deleterious or beneficial
mutations, in contrast, should fix less or more often, respectively. Time series
sequence data therefore suggest a simple way to investigate average properties
of different classes of mutations.

\paragraph{Synonymous polymorphisms in \env, C2-V5 are mostly deleterious}

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env, C2-V5, in patient 8\citep{shankarappa_consistent_1999,liu_selection_2006}.
Despite many synonymous mutations reaching high frequency, very few fix. This
observation in further quantified in panels \ref{fig:fixp1} and
\ref{fig:fixp2}, that stratify the data of
7-10 patients (see methods) according to the frequency at which different
mutations are observed. Considering all mutations in a frequency interval
$\nu_0$ at some time $t_i$, we calculate the fraction that is found at frequency
1, frequency 0, or at intermediate frequency at later time points $t_f$.
Plotting these fixed, lost, and polymorphic fraction against the time interval
$t_f-t_i$, we see that most synonymous mutations segregate for roughly one year
and are lost  much more frequently than expected. The ultimate probability of
loss or fixation is shown as a function of the initial frequency $\nu_0$ in
panel ??. In contrast to synonymous mutations, the non-synonymous seem to follow
more a less the neutral expectation -- a point to which we will come back below.

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency
 trajectories observed in patient 8~\cite{shankarappa_consistent_1999};
 Nonsynonymous and synonymous mutations are shown as solid and dashed lines,
 respectively. Colors indicate the position of the site along the C2-V5 region
 (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
 SEPARATELY. While non-synonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies.}
\label{fig:aft}
\end{center}
\end{figure}

\citet{bunnik_autologous_2008} present a longitudinal data sets on the entire
\env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
sequences each. Repeating the above analysis separately on the C2-V5 region
studied above and the remainder of \env~ reveal strikingly different behavior
inside and outside the hypervariable region. Within C2-V5, this data fully
confirms the observations made in the data set by
\citet{shankarappa_consistent_1999}. In the remainder of \env, however, observed
synonymous mutations behave as if they were neutral; see \FIG{fixp}.

@@

These observations suggest that many of the synonymous mutations in the part of
\env~that includes the hypervariable regions are deleterious, while outside
this regions mostly roughly neutral mutations are polymorphic.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}
\label{fig:fixp2}}
\caption{Left panel: time course of loss and fixation of synonymous mutations
 observed in a frequency interval $\nu_0$. The ultimate fraction of synonymous
 mutations that fix as a function of intermediate frequency $\nu_0$ is the
 fixation probability.  Right panel: fixation probability of derived synonymous
alleles is strongly suppressed in C2-V5 versus other parts of the {\it env}
gene, and of nonsynonymous ones. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

\paragraph{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA
secondary structures are relevant for HIV replication, mutations in nucleotides
involved in those base pairs are expected  to be deleterious and to revert
preferentially.  Many functionally important secondary structure elements have
been characterized, including  the  \rev{} response element (RRE) to enhance
nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
well studied case is the interaction between viral reverse transcriptase, viral
ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}. It has been suggested early on that parts of the viral
genome that has the potential to form stems is better conserved that the
remainder~\citep{forsdyke_reciprocal_1995}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPE} (left panel) in a cumulative histogram, the reactivity of
fixed alleles are systematically larger than of alleles that are doomed to
extinction. In other words, alleles that are likely to be breaking RNA helices
are also more likely to revert and finally be lost from the population. We then
split the synonymous mutations in the C2-V5 region further into conserved and
variable regions and found that the biggest depression in fixation probability
is observed in the conserved stems, while the variable loops show little
deviations from the neutral signature; see \FIG{SHAPE}B.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness effect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human housekeeping genes~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not find, however, evidence for any contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, although codon-optimized HIV seems to perform better
{\it in vitro}, the distance in CUB between HIV and human genes is not shrinking
at the macroevolutionary level. Second, within a single patient, we do not
observe any bias towards more human-like CUB in the synonymous mutations that
reach fixation rather than extinction. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}}
\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
in vitro} chemical attack and shown that some nucleotides are more likely to
be involved in RNA secondary folds. C1-C5 regions, in particular, show
conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
among all derived alleles in those regions reaching frequencies of order one,
there is a negative correlation between fixation and involvement in a base
pairing in a RNA stem (left panel). The rest of the genome does not show any
correlation (right panel). There might be too few silent polymorphisms in the
first place, or the signal might be masked by non-functional RNA
structures. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

\paragraph{Deleterious mutations are brought to high frequency by hitch-hiking}
While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~in which
we observe deleterious mutations at high frequency is special in that it
undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
recombination in HIV \cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency~\citep{smith_hitch-hiking_1974}.

The potential for hitch-hiking is already apparent from the allele frequency
trajectories in \FIG{aft}, where many mutations appear to change rapidly in
frequency as a flock. Deleterious synonymous mutations can be amplified
exponentially by selection on linked nonsynonymous sites, a process known as
{\it genetic draft}~\citep{gillespie_genetic_2000, neher_genetic_2011}. In order
to be advected to high frequency by a linked adaptive mutation, the deleterious
effect of the mutation has to be substantially smaller than the adaptive effect.
The latter was estimated to be on the order of $s_a = 0.01$ per day. The
approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp} (left panel), that shows the distribution of times for synonymous
alleles to reach the fix or get lost starting from intermediate frequencies. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects of
roughly $s_d \sim - 0.002$ per day.
